Words aren't created equal: Investigating bias on the CDI

### 2021-10-27 paper/cogsci.Rmd
adding refs, some edits and fixes

---
title: "Investigating Item Bias "
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

- what is our goal for measuring vocabulary? - identifying language delays, predicting later reading, or talking...
- how you select items influences how well you achieve these goals (and what bias you find)
We take the approach put forward by @Stenhaug2021: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.

@@

#Add this exclusion to the Analysis 
removed <- d_demo %>%
  filter(comprehension!=0)
#this is the actually d_demo we used so we must use this in our report
d_demo <- d_demo %>% 
  #filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
  arrange(data_id) %>%
  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
         ses_group = 
           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
```
We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2016; @Frank2021].

@@

```{r}
sex_tab <- table(d_demo$sex)
ses_tab <- table(d_demo$ses_group)
eth_tab <- table(d_demo$ethnicity)
eth2_tab <- table(d_demo$eth_group)
```

Research was conducted over  `r nrow(d_demo)` participants, whose data was parent reported and collected by the Wordbank Project. Unlike other projects Wordback relies on the kindness of others to contribute their data which means that often meta-data for some sets is missing. Even so, the project now has data over 29 different languages on children from ages 8-30 months old, though our study focused entirely on American English and Children 16-30 months.  When they received complete information, the Wordbank data set collected demographic data consisting of Birth Order, Race/Ethnicity, Sex and Mothers Education. For our study we focused on the comparing data between demographic groups on the axis of Sex, Ethnicity and Mothers Education, which was a proxy for social economic status. Out data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants, and an additional `r sum(is.na(d_demo$sex))` participants whose sex remained unreported. In terms of Ethnicity data was recorded from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` “Other” participants as well as `r sum(is.na(d_demo$ethnicity))` unreported. Given our distribution of ethnicities, for our analysis we split participants into White (`r eth2_tab["White"]`) and Non-White (`r eth2_tab["Nonwhite"]`). Finally, we split participants into high and low SES, of which we had `r ses_tab["high"]` high SES participants, corresponding to mothers' education of some college or higher and `r ses_tab["low"]` Low SES participants. 
<!-- same for eth, SES -->

<!-- goal: make custom glimmer plot in R with high / low bias words -->
```{r 2-col-image, fig.env = "figure*", fig.pos = "h", fig.width=4, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "This image spans both columns. And the caption text is limited to 0.8 of the width of the document."}
img <- png::readPNG("figs/walrus.png")
grid::grid.raster(img)
```

@@

Single column is the default option, but if you want set it explicitly, set `fig.env` to `figure`. Notice that the `num.cols` option for the caption width is set to `1`.

```{r image, fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=2, fig.height=2, set.cap.width=T, num.cols.cap=1, fig.cap = "One column image."}
img <- png::readPNG("figs/lab_logo_stanford.png")
grid::grid.raster(img)
```

## R Plots

You can use R chunks directly to plot graphs. And you can use latex floats in the
fig.pos chunk option to have more control over the location of your plot on the page. For more information on latex placement specifiers see **[here](https://en.wikibooks.org/wiki/LaTeX/Floats,_Figures_and_Captions)**

```{r plot, fig.env="figure", fig.pos = "H", fig.align = "center", fig.width=2, fig.height=2, fig.cap = "R plot" }
x <- 0:100
y <- 2 * (x + rnorm(length(x), sd = 3) + 3)

ggplot2::ggplot(data = data.frame(x, y), 
                aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

## Tables

Number tables consecutively; place the table number and title (in
10 point) above the table with one line space above the caption and
one line space below it, as in Table 1. You may float
tables to the top or bottom of a column, set wide tables across both
columns.

You can use the xtable function in the xtable package.

```{r xtable, results="asis"}
n <- 100
x <- rnorm(n)
y <- 2*x + rnorm(n)
out <- lm(y ~ x)

tab1 <- xtable::xtable(summary(out)$coef, digits=c(0, 2, 2, 1, 2), 
                       caption = "This table prints across one column.")

print(tab1, type="latex", comment = F, table.placement = "H")
```

>>>>

---
title: "Words aren't created equal: Investigating bias on the CDI"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

- what is our goal for measuring vocabulary? - identifying language delays, predicting later reading, or talking...
- how you select items influences how well you achieve these goals (and what bias you find)
We take the approach put forward by @stenhaug2021treading: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.

Differential item functioning (DIF) is a technique in IRT used to identify items that show bias against demographic groups. 
DIF is a statistical characteristic of an item that shows the extent to which the item might operate differently or measure varying abilities for subgroups and members of separate demographic groups. 
DIF is, however, a challenging metric for identifying bias for a couple reasons. 
Firstly, DIF is extremely hard to accurately root out. 
The process of finding DIF involves using the test you used to do IRT analysis in order to look for bias within the very same test. 
Secondly, the presence of DIF does not necessarily indicate the presence of bias; it can indicate that one demographic has a higher average ability level than the other. 
These two issues are the basis upon which we get the Fundamental DIF Identification problem.

This problem can be be understood through a concrete example that is well-known from research on early language learning: consider the fact that females show a larger vocabulary than males across early development [see @Frank2021 Ch. 6?; OTHER REFS]. 
The question however is whether or not girls actually have a higher ability level or if the tests have words that are bias toward girls ie. are easier for girls, or perhaps the test is biased toward boys but the natural ability of girls is enough to overcome that bias. To answer this question we need to know the ability levels of girls and boys in order to confirm that if a word is learned earlier by girls it is simply because of an ability difference rather than a bias inherent to the word. The problem arises from the fact that we measure ability level using the very same test that we are trying to check for biased words. If the boys and girls were of the same ability level this wouldn’t be an issue but given the evidence showing that girls have a higher ability level, we need to know the difference in ability so that when we find DIF for a specific word can be sure it is outside of the expecting DIF that results naturally from their difference in ability.

@@

#Add this exclusion to the Analysis 
removed <- d_demo %>%
  filter(comprehension!=0)
#this is the actually d_demo we used so we must use this in our report
d_demo <- d_demo %>% 
  #filter(comprehension!=0, !is.na(sex)) %>% # can't fit children not producing words, or with NA sex in group model
  arrange(data_id) %>%
  mutate(eth_group = ifelse(ethnicity=="White", "White", "Nonwhite"),
         ses_group = 
           ifelse(is.element(mom_ed, c("None", "Primary", "Some Secondary", "Secondary")), "low", "high"))
```
We analyze American English CDI: Words & Sentences administrations from `r nrow(d_demo)` children aged `r min(d_demo$age)` to `r max(d_demo$age)` from Wordbank [@Frank2017; @Frank2021].

@@

```{r, echo=F}
sex_tab <- table(d_demo$sex)
ses_tab <- table(d_demo$ses_group)
eth_tab <- table(d_demo$ethnicity)
eth2_tab <- table(d_demo$eth_group)
```

Research was conducted over  `r nrow(d_demo)` participants, whose data was parent reported and collected by the Wordbank Project. 
Unlike other projects Wordbank relies on the kindness of others to contribute their data which means that often meta-data for some sets is missing. 
Even so, the project now has data over 29 different languages on children from ages 8-30 months old, though our study focused entirely on American English and Children 16-30 months. 
When they received complete information, the Wordbank data set collected demographic data consisting of Birth Order, Race/Ethnicity, Sex and Mothers Education. 
We focused on comparing data between demographic groups on the axes of Sex, Ethnicity and Mother's Education, a proxy for socioeconomic status (henceforth, SES). 
Our data included `r sex_tab["Female"]` female and `r sex_tab["Male"]` male participants, and an additional `r sum(is.na(d_demo$sex))` participants whose sex remained unreported. 
In terms of Ethnicity data was recorded from `r eth_tab["White"]` white, `r eth_tab["Asian"]` Asian, `r eth_tab["Black"]` Black, `r eth_tab["Hispanic"]` Hispanic, and `r eth_tab["Other"]` “Other” participants as well as `r sum(is.na(d_demo$ethnicity))` unreported. Given our distribution of ethnicities, for our analysis we split participants into White (`r eth2_tab["White"]`) and Non-White (`r eth2_tab["Nonwhite"]`). 
Finally, we split participants into high and low SES, of which we had `r ses_tab["high"]` high SES participants, corresponding to mothers' education of some college or higher and `r ses_tab["low"]` Low SES participants. 
<!-- same for eth, SES -->

<!-- goal: make custom glimmer plot in R with high / low bias words -->
```{r 2-col-image, fig.env = "figure*", fig.pos = "h", fig.width=4, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "This image spans both columns. And the caption text is limited to 0.8 of the width of the document."}
#img <- png::readPNG("figs/walrus.png")
#grid::grid.raster(img)
```

@@

```{r image, fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=2, fig.height=2, set.cap.width=T, num.cols.cap=1, fig.cap = "One column image."}
#img <- png::readPNG("figs/lab_logo_stanford.png")
#grid::grid.raster(img)
```
